---
title: "ASV Affinity Abundance"
author: "Demi Glidden, Susan Collins, Natalie Winans"
date: "2/24/2021"
output: html_document
---

This document details data preparation for use in phyloseq. It uses metadata, Funguild output, DADA2 ASV table, and ASV affinities produced by custom logistic modeling R (not shown here).

Code is commented for clarity.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

## Reading in the data

```{r data, message=FALSE}
library(tidyverse)
library(dplyr)
library(phyloseq); packageVersion("phyloseq")
library(tibble)

setwd("/Users/demiglidden/bioinformatics/Panama/StatisticalAnalyses")

#read in data
ids <- read_csv("ASV_ID_seq.csv")
classified <- read.csv("classified_ASVs.csv")
funguild <- read_tsv("transformed_ASVfunguildtable.guilds.txt")
metadata <- read_tsv("panama_metadata_full_v1.txt")
ASV_10574 <- read_tsv("PanamaPrecip_ITS_ASVs_r10574.txt")
```

## Data preparation

Rename columns for consistency:

```{r dataprep}
#rename sequence columns so they match (ASV to ASV_ID)
classified <- classified %>% dplyr::rename(ASV_ID = ASV)
#rename to ASV_seq
funguild <- funguild %>% dplyr::rename(ASV_seq = otuID)
#rename X to ASV_seq
ASV_10574 <- ASV_10574 %>% dplyr::rename(ASV_seq = X)
#drop sample counts
affinities <- classified %>% select(c(Affinity, ASV_ID))
```

The data frames need to merged with the ASV_IDs, then we can begin to make data frames that can be turned into phyloseq objects.

```{r mergingfiles}
#join the two data frames so that the affinities have ASV_IDs and sequences
affinity_ids <- left_join(affinities, ids, by = "ASV_ID")
affinity_ids <- left_join(affinity_ids, funguild, by = "ASV_seq")

###making taxa and count tables to turn them into phyloseq objects

#get preliminary ASV count table
asv_counts <- classified %>% dplyr::select(2:347)
#remove duplicates 
asv_counts <- asv_counts %>% distinct(ASV_ID, .keep_all=TRUE)
#add seqs for use as row names
asv_counts <- merge(asv_counts, ids, by="ASV_ID")
#move seqs to front of data frame (not necessary, but tidy)
asv_counts <- asv_counts %>% relocate(ASV_seq, .after = ASV_ID)
```

Moving forward with making phyloseq-worthy objects:

```{r phyloseqprep}
#get prelim taxa table
asv_taxa <- ASV_10574 %>% dplyr::select(1:8)
#merge with affinity IDs
##the following steps also remove ASVs that do not have affinities, using the "right_join"
asv_taxa <- right_join(asv_taxa, affinity_ids, by ="ASV_seq")
asv_taxa <- as.data.frame(asv_taxa)
#drop sequences, this part is opttional if you don't want them in your final phyloseq object
#asv_taxa <- asv_taxa %>% dplyr::select(!c("ASV_seq"))
#remove duplicates 
asv_taxa <- asv_taxa %>% distinct(ASV_ID, .keep_all=TRUE)

## make tables with only ASV IDs that are present in taxa and count tables
common_ASVs <- merge(asv_taxa,asv_counts,by='ASV_seq')

# separate the tables again
#this will pull out the counts
counts <- common_ASVs %>% dplyr::select(contains("_"))
#this will pull out the taxonomy and affinities. if you want more info from the metadata, you can remove or add to the selection below with column numbers
taxa <- common_ASVs %>% dplyr::select(1:10,12,14,15)

#clean up after the merge: go back to one column of ASV_IDs, and rename them as such
counts <- counts %>% select(-ASV_ID.x)
counts <- counts %>% dplyr::rename(ASV_ID = ASV_ID.y)
taxa <- taxa %>% dplyr::rename(ASV_ID = ASV_ID.x)

#make ASV_IDs rownames
taxa <- tibble::column_to_rownames(taxa, 'ASV_seq')
taxa <- as.matrix(taxa)
#make ASV_IDs rownames
counts <- tibble::column_to_rownames(counts, 'ASV_seq')
counts[] <- lapply(counts, as.numeric)
```
  
## Making Phyloseq Objects

All of the above transformations prepare the data to be read in by phyloseq and made into phyloseq objects.
  
```{r phyloseq}
#make phyloseq objects
counts_phylo = otu_table(counts, taxa_are_rows = TRUE)

taxa_phylo = tax_table(taxa)

ps <- phyloseq(counts_phylo,taxa_phylo)
```
  
A simple plot showing abundances of ASVs with affinities for precipitation and phosphorous.
  
```{r plot}
plot_bar(ps, x="Affinity") + 
  theme_classic() +
  ggtitle("Abundance of ASVs with MAP and P Affinities")
```